---
title: "Model Fitting"
output:
  rmarkdown::html_vignette: default
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{fontspec}
  - \usepackage{etoolbox}
  - \usepackage{fvextra}
  - \usepackage{needspace}
  - \definecolor{shadecolor}{RGB}{232,232,232}
  - \fvset{breaklines=true,breakanywhere=true}
  - \BeforeBeginEnvironment{Shaded}{\Needspace{10\baselineskip}}
  - \BeforeBeginEnvironment{Shaded}{\vspace{0.5em}}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \RecustomVerbatimEnvironment{verbatim}{Verbatim}{breaklines=true,breakanywhere=true}
css: styles.css
vignette: >
  %\VignetteIndexEntry{Model Fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-defaults, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(
    fig.dim=c(5, 5), fig.show="hold", out.width="50%",
    echo=TRUE, message=FALSE, warning=FALSE
)
```

This article shows how to turn raw one-dimensional NMR spectra into a trained
classifier with `metabodeconplus`, following the same pipeline described in the
paper:

1. **Deconvolution** -- represent each spectrum as a list of Lorentzian peaks.
2. **Alignment (CluPA)** -- shift peaks so corresponding peaks across spectra
   share the same chemical-shift index.
3. **Peak snapping (RefPA)** -- snap each aligned peak onto the nearest
   reference-grid column, so all spectra share one common set of feature
   columns.
4. **Feature matrix** -- collapse the snapped peak lists into one row per
   spectrum.
5. **Classification** -- fit a random forest with
   [`ranger`](https://cran.r-project.org/package=ranger).

`fit_mdm()` runs all five steps in one call; below we first do them by hand so
each intermediate state is visible, and then reproduce the result with the
one-shot call. We use the bundled `sim2` dataset.

# Load spectra

`sim2` contains 100 simulated 1D NMR spectra split evenly into groups `A` and
`B`. Five of every 25 peaks per spectrum differ between groups by 10 % in area.
The group labels are attached as an attribute; see `?sim2` for details.

```{r load-spectra}
library(metabodeconplus)
# Small rank-based AUC helper (positive class = second factor level).
auc <- function(y, prob) {
    pos <- y == levels(y)[2]; r <- rank(prob)
    n1 <- sum(pos); n0 <- sum(!pos)
    if (n1 == 0 || n0 == 0) NA_real_ else (sum(r[pos]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
x <- sim2
y <- attr(sim2, "group")
n <- length(x)
```

We use one half of the data for training and one half for testing.

```{r split}
set.seed(1)
tr <- sort(sample(n, round(0.5 * n)))
te <- setdiff(seq_len(n), tr)
true_x0 <- attr(sim2, "true_x0")   # ppm of the discriminating peaks
```

# The pipeline, step by step

We pick a small group-balanced subset (4 + 4 spectra) just for the plots; the
model itself is trained on all of `x[tr]`.

```{r prep}
abtr <- c(which(y[tr] == "A")[1:4], which(y[tr] == "B")[1:4])
yab <- y[tr][abtr]
```

## Step 1: Deconvolute

`deconvolute()` models each spectrum as a superposition of Lorentzian peaks.

```{r decon}
decons <- deconvolute(x[tr], nfit=10, smit=2, smws=5, delta=10, npmax=0, verbose=FALSE)
plot_spectra(decons[abtr])
heat_spectra(decons[abtr], y=yab)
```

## Step 2: Align (CluPA)

`clupa()` shifts each spectrum's peaks toward a reference spectrum using the
hierarchical cluster-based peak alignment (CluPA) algorithm. It picks the
reference automatically and attaches it, so we can reuse it for the test data.

```{r align}
aligns <- clupa(decons, maxShift=50, verbose=FALSE)
ref <- attr(aligns, "ref")
plot_spectra(aligns[abtr])
```

## Step 3: Snap peaks to the reference (RefPA)

CluPA aligns peaks continuously; `snap_to_ref()` then snaps each peak onto the
nearest reference-grid column (within `maxCombine` datapoints), so every
spectrum ends up described by the *same* set of feature columns.

```{r snap}
snapped <- snap_to_ref(aligns, maxCombine=5)
```

## Step 4: Build the feature matrix

`peak_mat()` rasterises the snapped peak lists into a matrix with one row per
spectrum and one column per populated reference-grid position. `peakPos`
records which columns are populated -- these are the features the model sees.

```{r featmat}
X <- peak_mat(snapped)
peakPos <- attr(X, "peakPos")
dim(X)
heat_spectra(X, y=y[tr], true_x0=true_x0)
heat_spectra(X, y=y[tr], true_x0=true_x0, scale_cols=TRUE)
```

The standardized view (`scale_cols=TRUE`) makes the group structure visible:
columns near `true_x0` show consistent sign differences between A (top) and B
(bottom).

## Step 5: Fit a random forest

We fit a probability random forest on the feature matrix with `ranger` and read
off the out-of-bag (OOB) error.

```{r ranger, eval=requireNamespace("ranger", quietly=TRUE)}
rf <- ranger::ranger(x=X, y=y[tr], probability=TRUE, num.trees=500, seed=1)
cat(sprintf("OOB error: %.1f%%\n", 100 * rf$prediction.error))
```

# The one-shot call

`fit_mdm()` performs all five steps -- deconvolute, align, snap, featurize,
fit -- and, when any of `npmax` / `maxShift` / `maxCombine` is a vector,
searches the grid and returns the best model. Choose the backend with
`model = "ranger"`.

```{r fit-mdm, eval=requireNamespace("ranger", quietly=TRUE)}
md <- fit_mdm(x[tr], y[tr], model="ranger",
              npmax=0L, maxShift=50L, maxCombine=5L,
              verbosity=0, nworkers=1)
print(md)
```

# Predict held-out spectra

`predict()` on an `mdm` object mirrors the training pipeline for new data,
reusing the stored reference and feature columns.

```{r predict, eval=requireNamespace("ranger", quietly=TRUE)}
preds <- predict(md, x[te], type="all", verbosity=0)
acc <- mean(preds$class == y[te])
au  <- auc(y[te], preds$prob)
cat(sprintf("Test accuracy: %.1f%%\n", 100 * acc))
cat(sprintf("Test AUC:      %.3f\n", au))
```

# Tune the preprocessing

Passing a vector for `npmax`, `maxShift` or `maxCombine` makes `fit_mdm()`
evaluate the cartesian product and keep the best-scoring cell (accuracy, ties
broken by AUC). The augmented grid is returned in `md$mog`.

```{r tune, eval=requireNamespace("ranger", quietly=TRUE)}
mt <- fit_mdm(x[tr], y[tr], model="ranger",
              npmax=c(0L, 30L), maxShift=c(20L, 50L), maxCombine=c(2L, 5L),
              verbosity=0, nworkers=1)
knitr::kable(head(mt$mog[order(-mt$mog$auc), ], 5), row.names=FALSE,
             caption="Top parameter combinations by AUC.")
```

For an honest generalization estimate on small datasets, wrap the whole search
in outer cross-validation with `benchmark()` (not run here because it repeats
the grid search for every fold):

```{r benchmark, eval=FALSE}
bm <- benchmark(x, y, model="ranger", npmax=0L, maxShift=50L, maxCombine=5L, k=5)
mean(bm$predictions$true == bm$predictions$pred)
```
